Weakly Nonlinear Analysis of the Standard and Helical Magnetorotational Instabilities in a Taylor Couette Flow

### 2016-07-16 paper/widegap_mri_paper.tex
high res run needs more time

Because of the MRI's complexity, analytical treatments of the MRI use some combination of convenient approximations: ideal MHD, linearized equations, expedient boundary conditions, and simplified geometries. In particular, much of the analytical work on MRI saturation has focused on local approximations, often representing a section of a disk by solving the MHD equations in an isolated region subject to shear periodic boundary conditions (the ``shearing box"). While the MRI is a local instability, certain properties of the local problem are not generic to the global problem. Channel modes, a class of MRI mode that dominates MRI evolution and saturation in the local approximation, are artificially over-prominent in the shearing box (e.g. \citei{Latter:2015}).

@@

Because we include curvature terms, our theory is also relevant to the helical magnetorotational instability (HMRI). Discovered by \citeauthor{Hollerbach:2005tr} (\citeyear{Hollerbach:2005tr}), the HMRI is an overstability in which the background magnetic field is helical. The HMRI currently occupies a special place in the MRI puzzle. The HMRI has been proposed as a method of awakening angular momentum transport in the ``dead zones" of protoplanetary disks where the magnetic Prandtl number ($Pm = \nu/\eta$) becomes very small. However the rotation profiles needed to excite HMRI may be prohibitively steeper than Keplerian depending on the boundary conditions, and so its role in astrophysical disks is currently debated (\citei{Liu:2006}, \citei{Rudiger:2007}, \citei{Kirillov:2013}). However, the HMRI is significantly easier to excite in a laboratory setting than the standard MRI, and has already been detected in the laboratory by the Potsdam Rossendorf Magnetic Instability Experiment (PROMISE; \citei{Stefani:2006iv}, \citei{Stefani:2009hp}).

>>>>

Because of the MRI's complexity, analytical treatments of the MRI use some combination of convenient approximations: ideal MHD, linearized equations, expedient boundary conditions, and simplified geometries. In particular, much of the analytical work on MRI saturation has focused on local approximations, often representing a section of a disk by solving the MHD equations in an isolated region subject to shear periodic boundary conditions (the ``shearing box"). While the MRI is a local instability, certain properties of the local problem are not generic to the global problem. In the shearing box, linear evolution is dominated by a class of MRI mode known as channel modes.  %Channel modes, a class of MRI mode that dominates MRI evolution and saturation in the local approximation, are artificially over-prominent in the shearing box (e.g. \citei{Latter:2015}).

@@

Because we include curvature terms, our theory is also relevant to the helical magnetorotational instability (HMRI). Discovered by \citeauthor{Hollerbach:2005tr} (\citeyear{Hollerbach:2005tr}), the HMRI is an overstability in which the background magnetic field is helical. The HMRI currently occupies a special place in the MRI puzzle. The HMRI has been proposed as a method of awakening angular momentum transport in the ``dead zones" of protoplanetary disks where the magnetic Prandtl number ($Pm = \nu/\eta$) becomes very small. However the rotation profiles needed to excite HMRI may be prohibitively steeper than Keplerian depending on the boundary conditions, and so its role in astrophysical disks is currently debated (\citei{Liu:2006}, \citei{Rudiger:2007}, \citei{Kirillov:2013}). The HMRI is significantly easier to excite in a laboratory setting than the standard MRI, and has already been detected in the laboratory by the Potsdam Rossendorf Magnetic Instability Experiment (PROMISE; \citei{Stefani:2006iv}, \citei{Stefani:2009hp}). The standard MRI has yet to be detected in the laboratory, although several plasma experiments are making progress.
